---
title: "Industrial Impact on Housing Prices in Western Kentucky"
author: "Brad Brauser"
date: "6/25/2023"
output: 
word_document: default
pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(knitr)
library(formattable)
library(ggplot2)
library(tidyr)
library(geosphere)
library(Rcpp)
library(kableExtra)
library(stargazer)
```

## Summary

## Introduction

In the western Kentucky area, the impact of proximity to industrial areas on property pricing is a significant concern. This project aims to examine how residential properties' value is affected by their proximity to factories, packaging facilities, and other industries. By analyzing factors such as distance, pollution levels, and market dynamics, we seek to provide valuable insights for homeowners, real estate agents, and policymakers. Understanding the correlation between industrial proximity and property prices in this region is crucial for informed decision-making regarding real estate investments and urban planning.

## Literature Review

Although there is limited specific literature on the impact of industrial areas on property pricing in western Kentucky, studies conducted in other regions shed light on similar issues. One such study examined the property prices in Bordeaux, Dunkirk, and Rouen, which have hazardous chemical and petrochemical industries in or near the cities (Grislain-Letremy & Katossky, 2014). Another study focused on regions in the Netherlands, namely Rancid and Noord-Brabant (F. Vor & H. De Groot, 2011). Both papers found that the influence of industrial areas on property prices is predominantly localized, with the presence and size of industrial sites affecting the perceived spatial quality of the neighborhood. Furthermore, the willingness to pay for prevention measures varied significantly among different industrial areas (Grislain-Letremy & Katossky, 2014). While these studies offer valuable insights, there is a need for further research specifically addressing the western Kentucky area to understand the localized impact of industrial proximity on property pricing and the unique dynamics of the region.

Existing studies have examined the impact of industrial areas on property pricing in various regions, yet there remains a significant gap in research specifically addressing the western Kentucky area. This region possesses unique characteristics, including a distinct industrial landscape and residential patterns, which necessitate a dedicated study. Conducting research in this problem area will address the lack of localized understanding and insights into how proximity to industrial areas affects property values in western Kentucky. The study aims to investigate the direct correlation between distance from industrial sites and property pricing, the influence of pollution levels on housing market dynamics, and the varying perceptions of spatial quality in neighborhoods adjacent to different industrial sectors. By filling this research gap, the study will provide homeowners, real estate agents, and policymakers with essential information to make informed decisions regarding property investments, urban planning, and environmental management in the western Kentucky area. Ultimately, it will contribute to a comprehensive understanding of the impact of industrial proximity on property pricing and help shape strategies for sustainable development and equitable growth in the region.

Understanding the correlation between these factors is vital for homeowners, real estate agents, and policymakers who make decisions about property investments and urban planning. By investigating the proximity of residential areas to industrial sites, the study aims to reveal whether there is a direct correlation between distance and housing prices. This information will provide valuable insights into the dynamics of the housing market in Western KY, allowing stakeholders to make informed decisions based on the spatial relationship between industrial areas and property values. Additionally, the study's findings will shed light on the potential impacts of industrial activities, such as pollution levels, on housing prices. This knowledge is essential for policymakers in developing strategies to ensure the sustainable development of the region while safeguarding the welfare of residents. By exploring these research questions, the study aims to contribute to evidence-based decision-making and support the development of a thriving real estate market in Western KY.

## Theory

This paper is exploring the impact between an industrial area's location and the pricing surrounding residential properties. This exploration includes:

```         
H1. The closer that a residential property is to an industrial area, the lower in price the property will be.
```

## Data

Housing data was downloaded in a CSV from <https://westky.paragonrels.com/ParagonLS/Default.mvc#1,1,2> and uploaded to <https://github.com/bbrauser/WKYIndustrialImpact/tree/main/Data>. The data includes:

```{r}

# Specify the URL of the CSV file
housing_url <- "https://raw.githubusercontent.com/bbrauser/WKYIndustrialImpact/main/HousingResearch.csv"

# Read the CSV file from the URL
housing_df <- read.csv(url(housing_url))

# Display the first 6 rows of the df
head(housing_df)
```

Because some of these properties were either not sold or are rental properties, those rows can be removed. The following is a list of the values removed and the reason why they were removed:

  Withdrawn: The property did not sell
  ContractReceived: The property is under contract, but the price it was closed for is unknown
  Expired: The property did not sell
  Rented: The property was a rental
  Auction: The property was auctioned off and not sold
  
```{r}
# Define the values to be removed
values_to_remove <- c("Withdrawn", "Contract Received", "Expired", "Rented", "Auction")

# Remove the values from the Status column
housing_df <- subset(housing_df, !(Status %in% values_to_remove))

# Reset the index of the dataframe
housing_df <- data.frame(housing_df, row.names = NULL)

# Display the filtered dataframe
head(housing_df)
```

Some rows did not have values in Latitude and Longitude, so those were removed:

```{r}
# Remove rows with blank cells in Latitude and Longitude columns
housing_df <- housing_df[complete.cases(housing_df$Latitude, housing_df$Longitude), ]

# Display the filtered dataframe
head(housing_df)
```
To give more of a clear minimum price, the Price column was filtered to show only prices of $20,000 and above:

```{r}

# Filter the price column to show values 20,000 and above
housing_df <- filter(housing_df, Price >= 20000)

# Display the first 6 rows of the df
head(housing_df)
```

Some of the data was not needed, so the housing data was limited to Address, County, Price, Latitude, and Longitude:

```{r}

# Removes unneeded columns from df
housing_df <- select(housing_df, Address, County, Price, Latitude, Longitude)

# Display the first 6 rows of the df
head(housing_df)
```

Location information of the industrial areas in western Kentucky was downloaded in a CSV from <https://www.irs.gov/pub/irs-soi/17incyky.xlsx> and uplaoded to GitHub. The data includes:

```{r}

# Specify the URL of the CSV file
industry_url <- "https://raw.githubusercontent.com/bbrauser/WKYIndustrialImpact/main/IndustrialAreas.csv"

# Read the CSV file from the URL
industry_df <- read.csv(url(industry_url))

# Display the first 6 rows of the df
head(industry_df)
```

This data only needs County, FacilityNa, Longitude, and Latitude:

```{r}

# Compiles dataframe from industry_df
industry_df <- select(industry_df, County, FacilityNa, Longitude, Latitude)

# Display the first 6 rows of the df
head(industry_df)
```

Since this df includes industrial locations in the whole state of Kentucky, the data must be filtered to include only the counties that are present in the houisng data CSV:

```{r}

industry_df <- industry_df[industry_df$County %in% housing_df$County, ]

# Resets the index
industry_df <- data.frame(industry_df, row.names = NULL)

# Display the first 6 rows of the df
head(industry_df)
```

## Methodology

After cleaning and preparing the data for analysis, a new dataframe ('price_analysis_df') was created that concatenated

```{r}
# Initialize an empty data frame to store the results
price_analysis_df <- data.frame(Address = housing_df$Address,  
                               Price = housing_df$Price,  
                               stringsAsFactors = FALSE)  

# Add columns for closest industry and distance  
price_analysis_df$Industry <- ""  
price_analysis_df$Distance <- Inf  

# Calculate distance between each property and industry 
for (property_row in 1:nrow(housing_df)) {
  property_lat <- housing_df$Latitude[property_row]
  property_lon <- housing_df$Longitude[property_row]
  
  for (industry_row in 1:nrow(industry_df)) {
    industry_lat <- industry_df$Latitude[industry_row]
    industry_lon <- industry_df$Longitude[industry_row]
    
    distance <- round(sqrt((property_lat - industry_lat)^2 + (property_lon - industry_lon)^2), 2)
    
    # Check if the current distance is smaller than the previously recorded minimum distance
    if (distance < price_analysis_df$Distance[property_row]) {  
      price_analysis_df$Industry[property_row] <- industry_df$FacilityNa[industry_row]  
      price_analysis_df$Distance[property_row] <- distance
    }
  }
}  

# Print the resulting data frame  
head(price_analysis_df)
```

Because some of these properties are not very close to an industry, we are going to filter out this dataframe to only include properties that are 20 miles away or less from an industry in industry_df. This still gives us about 8000 properties to analyze:

```{r}
# Filter the price column to show values 20,000 and above
price_analysis_df <- filter(price_analysis_df, Distance <= 20)

# Display the first 6 rows of the df
head(price_analysis_df)

```

Below is a simple scatterplot that shows the relationship between Price and Distance:

```{r}

# Get the maximum values 
max_price <- max(price_analysis_df$Price)
max_dist <- max(price_analysis_df$Distance)

# Scatterplot of Price vs Distance
ggplot(price_analysis_df, aes(x = Price, y = Distance)) + 
  geom_point(color = "blue", size = 3) + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(
    title = "Price vs Distance",      
    x = "Price (USD)",               
    y = "Distance (miles)"           
  ) +
  theme_bw() +                       
  theme(
    plot.title = element_text(size = 15, face = "bold"),   
    axis.title = element_text(size = 12)                   
  ) +
  scale_x_continuous(breaks = seq(0, max_price, 75000), limits = c(0, max_price), labels = comma) +  
  scale_y_continuous(breaks = seq(0, max_dist, 1), limits = c(0, max_dist)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  

```

The residual plot below shows if the residuals are randomly distributed, which is an assumption of linear regression:

```{r}
# Fit a linear model 
lm <- lm(Price ~ Distance, data = price_analysis_df)

# Extract the residuals 
residuals <- residuals(lm)

# Get the maximum values 
max_price <- max(price_analysis_df$Price)
max_dist <- max(price_analysis_df$Distance)

# Scatterplot of Price vs Distance, colored by residuals  
ggplot(price_analysis_df, aes(x = Price, y = Distance, color = residuals)) + 
  geom_point(size = 3) + 
  geom_smooth(method = "lm", se = FALSE) + 
  scale_color_gradient2(low = "blue", high = "red", mid = "white") +
  labs(
    title = "Price vs Distance",      
    subtitle = "Colored by Residuals",
    x = "Price (USD)",               
    y = "Distance (miles)"           
  ) +
  theme_bw() +                       
  theme(
    plot.title = element_text(size = 15, face = "bold"),
    plot.subtitle = element_text(size = 12, face = "italic"),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 12)
  ) +
  scale_x_continuous(breaks = seq(0, max_price, 75000), limits = c(0, max_price), labels = comma) +  
  scale_y_continuous(breaks = seq(0, max_dist, 1), limits = c(0, max_dist)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

The quantile-quantile (or QQ) plot below shows the quantiles of the residuals to the quantiles of a normal distribution, allowing one to check if the residuals are normally distributed, which is another assumption of linear regression. Deviations from the diagonal line indicate non-normality.
```{r}
# Create the QQ plot
ggplot(price_analysis_df, aes(sample = Distance)) +
  stat_qq() +
  geom_abline(color = "red") +
  xlab("Theoretical Quantiles") +
  ylab("Sample Quantiles") +
  ggtitle("QQ Plot of Distance")

ggplot(price_analysis_df, aes(sample = Price)) +
  stat_qq() +
  geom_abline(color = "red") +
  xlab("Theoretical Quantiles") +
  ylab("Sample Quantiles") +
  ggtitle("QQ Plot of Price")
```

 ## Results
 
 The plots shown above give a visual representation of the relationship between Price and Distance which seems to indicate that there is a negative relationship between Price and Distance. That is, when Distance increases, Price decreases. To give a more granualr look at the relationship, a regressions table was created:


```{r}
# Fit linear regression model
model <- lm(Price ~ Distance, data = price_analysis_df)

# Create regression table
regression_table <- stargazer(model, 
                             title = "Regression Results",
                             dep.var.labels = "Price",
                             covariate.labels = "Distance",
                             header = FALSE,
                             type = "text",
                             digits = 3)

# Print regression table
cat(regression_table, sep = "\n")
```

The results indicated a significant negative effect of Distance on Price (-1,753.965, p < 0.01), suggesting that as the distance increased, the price decreased. However, there was some uncertainty in the estimate (standard error: 225.344).

The constant term in the regression equation represented 170,811.700 (p < 0.01), which was the expected price when the distance was zero. The R-squared value was 0.007, indicating that only 0.7% of the variation in Price was explained by Distance. The model's predictive power was limited, as seen from the low R-squared value. The residual standard error was 113,127.900, indicating the average prediction error.

Overall, the findings suggested that Distance had a significant but weak influence on Price.

## Conclusion and Implications

The initial theory that the closer to an industrial area a property is the more negatively the price is affected. This was proven not to be the case, as the closer a property was to an industrial area, the higher the price of the property was. While this study was limited to the western KY area, it might be beneficial to see the implications of not only the entire state of Kentucky, but also how the type of industry affects the price.

# References

de Vor, F., & de Groot, H. L. (2011). The impact of industrial sites on residential property values:
A hedonic pricing analysis from the Netherlands. Regional Studies, 45(5), 609–623.
https://doi.org/10.1080/00343401003601925

Grislain-Letrrmy, Cc., & Katossky, A. (2014). The impact of hazardous industrial facilities on
housing prices: A comparison of parametric and semiparametric hedonic price models.
SSRN Electronic Journal. https://doi.org/10.2139/ssrn.2457195

Team kentucky. Team Kentucky | Cabinet for Economic Development. (n.d.). https://ced.ky.gov/
existing_industries/Major_Industries
